### 2021-07-09 rad-icfp-2021/complexity.tex
fixing map2
minor correctness section

There is a first restriction that prevents our transformation from satisfying the cheap gradient principle.
If we try to show that the cheap gradient principle holds by induction on the term, it fails on the !map2!.
When differentiating !map2!, there are two series of calls to (sub)gradients of !e$_1$!. 
The induction hypothesis is then too weak to conclude. 
The problem comes from the fact that !e$_1$! could itself use !map2!. 
So the constant $K$ can be independant of $n$ and of 
the size of the term if we allow it to be dependent on the level of nestings of !map2!.
A similar phenomenon happens with !reduce!.

@@

\begin{example}
    As shown in appendix \ref{sub:gradintro}, the optimizations from Figure~\ref{fig:optim} 
    are sufficient to show that the gradients of the terms of the introduction reduce to the following.

>>>>

There is a first restriction that prevents our transformation from satisfying the cheap gradient principle.
If we try to show that the cheap gradient principle holds by induction on the term, it fails on the !map2!.
When differentiating !map2!, there are three series of calls to (sub)gradients of !e$_1$!. 
The induction hypothesis is then too weak to conclude. 
The problem comes from the fact that !e$_1$! could itself use !map2!. 
So the constant $K$ can be independant of $n$ and of 
the size of the term if we allow it to be dependent on the level of nestings of !map2!.
A similar phenomenon happens with !reduce!.

@@

\begin{example}
    As shown in the supplementary material,  
    %appendix \ref{sub:gradintro}, 
    the optimizations from Figure~\ref{fig:optim} 
    are sufficient to show that the gradients of the terms of the introduction reduce to the following.

### 2021-07-09 rad-icfp-2021/correctness.tex
minor correctness section

 \begin{theorem}
     For every $\Gamma \vdash$ !e:!$\reals$, we have 
     $\sem{\grad_\Gamma e}= \grad_\Gamma\sem{e}$ in the category of diffeological spaces.
 \end{theorem}

@@

A term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ is interpreted as a morphism $\sem{e}:[\sem{A_1},\ldots,\sem{A_n}]\to \sem{A}$ in a multicategory.
Substitution is interpreted as composition. We first consider a syntactic model for a language, which consists of a 
free multicategory on some base types and primitives. Our source and target languages induce syntactic multicategories as follows.

@@

We similarly define $\SynTarget$, the syntactic multicategory for Target.

@@

    \begin{tabular}{r c l} 
    !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>;swap;$(\seml$ e$_1\semr)^{\times n}$!
    \end{tabular}

where $\Delta_n$ is the $n$-copy map !<id,$\ldots$,id>!.
The semantics for !reduce (x,y.e$_1$) e$_2$ e$_3$! is given by first !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>!, following a swap isomorphism  $\Gamma^n \times \RR \times \RR^n \to \reals \times (\reals \times \Gamma)^n$. 
Finally, we apply $\seml$ e$_1\semr \times id$ $n$ times, where the identity is of the obvious type at each stage.

Variables are interpreted as projections $\pi_i$, let binding as composition in the multicategory. 
Pairs are interpreted using the Cartesian structure of the multicategory.
\end{definition}

We can similarly interpret Target in a multicategory of smooth-like spaces and functions. 
However, Target is higher-order and $\CartSp$ is not Cartesian Closed. 
Instead, we can interpret Target in the category of Diffeological spaces, 
as in \cite{huot2020correctness}. 
Diffeological spaces (\cite{iglesias2013diffeology}) are a conservative extension of $\CartSp$. 
The key idea will be that a higher-order function is called smooth if it sends smooth functions to smooth functions, meaning that we can never use it
to build first order functions that are not smooth.

@@

One main reason for introducing UNF is to have a better handle over
the computation flow of the term, similarly to what the ANF or CPS transformations do.
A convenient categorical setting for this is to use string diagrams.
To better fit the standard denotational semantics of languages, 
we can instead use concategories.

Concategories are a generalization of categories and multicategories in that 
they allow morphisms both from and to a list of objects. 
Similarly to categories and multicategories, 
most categorical construtions used for the semantics of functional languages 
have equivalent in concategories.

In particular, one can form a syntactic concategory on some base types and primitives, 
which satisfies a similar universal property as syntactic multicategories.
Two particular examples are of interest for us, as they will allow us to interpret Source and Target UNF
but they will also help us explain the UNF and $UNF^{-1}$ transformations.

@@

This is proved in Appendix \ref{sub:Reverse derivative of array operations}.

>>>>

@@

A term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ is interpreted in a multicategory as a morphism $\sem{e}:[\sem{A_1},\ldots,\sem{A_n}]\to \sem{A}$.
Substitution is interpreted as composition. We first consider a syntactic model for a language, which consists of a 
free multicategory on some base types and primitives. Our source and target languages induce syntactic multicategories as follows.

@@

We similarly define $\SynTarget$, the syntactic multicategory on the target language.

@@

Variables are interpreted as projections $\pi_i$, let binding as composition in the multicategory. 
Pairs are interpreted using the Cartesian structure of the multicategory.
This interpretation extends to !map2 (x,y.e$_1$) e$_2$ e$_3$!. It is given by
!<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>;swap;$(\seml$ e$_1\semr)^{\times n}$!
where $\Delta_n$ is the $n$-copy map !<id,$\ldots$,id>!, and !swap! a permutation.
Similarly, the semantics for !reduce (x,y.e$_1$) e$_2$ e$_3$! is then given by first !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>!, followed by a perumutation $\Gamma^n \times \RR \times \RR^n \to \reals \times (\reals \times \Gamma)^n$. 
Finally, we apply $\seml$ e$_1\semr \times id$ $n$ times, where the identity is of the obvious type at each stage.
\end{definition}

We can similarly interpret the target language in a multicategory of smooth-like spaces and functions. 
However, Target is higher-order and $\CartSp$ is not Cartesian Closed. 
Instead, we can interpret Target in the category of Diffeological spaces, 
as in \cite{huot2020correctness}. 
Diffeological spaces (\cite{iglesias2013diffeology}) are a conservative extension of $\CartSp$. 
The key idea will be that a higher-order function is called smooth if it sends smooth functions to smooth functions, meaning that we can never use it
to build first order functions that are not smooth.

@@

One main reason for introducing UNF is to have a better handle over
the computation flow of the term, in the same vein as the ANF or CPS transformations.
A convenient categorical setting for this is to use string diagrams.
To better fit the standard denotational semantics of languages, 
we use concategories instead. Similarly to categories and multicategories, 
most categorical constructions used for the semantics of functional languages 
have equivalent in concategories.

In particular, one can form a syntactic concategory on some base types and primitives, 
which satisfies a similar universal property as syntactic multicategories.
Two particular examples are of interest to us, as they will allow us to interpret Source UNF and Target UNF
but they will also help us explain the $\UNFSymbol$ and $UNF^{-1}$ transformations.

@@

This is proved in the supplementary material.
%Appendix \ref{sub:Reverse derivative of array operations}.
